A statistical assessment of influenza intensity thresholds from the moving epidemic and WHO methods

### 2024-06-03 Draft/Draft_MEM.tex
Major update of draft.

\documentclass{article}
\usepackage[left=2.5cm,right=2.5cm, top = 2.5cm, bottom = 3cm]{geometry}
\usepackage{graphicx}
\usepackage{natbib}
\usepackage{hyperref}
\usepackage{todonotes}
\usepackage{booktabs}
\usepackage{amssymb}
\usepackage{amsmath}

@@

@@

\begin{center}
\textbf{This manuscript is is a preprint and has not been subject to peer review.}
\end{center}

\abstract{
\begin{quote}
\footnotesize Intensity thresholds serve to make influenza activity comparable across countries and surveillance systems. The moving epidemic method (MEM) and the WHO method are widely used to classify season peaks as low, medium, high or very high. In both methods, thresholds correspond to quantiles of a normal distribution fitted to a set of potentially transformed historical observations. However, there are differences in how this set is constructed. We assess the impact of these choices analytically and in simulation studies. We find that under the default settings of both the MEM and WHO method, on average more season peaks than intended are classified as high or very high. Combining characteristics of both, better-calibrated thresholds can be achieved. Even these, however, must be expected to have rather modest sensitivity and positive predictive values.
\end{quote}
}

\begin{center}
\textbf{Extended summary}
\end{center}

{\begin{quote}
\footnotesize Intensity thresholds are a common tool to make influenza activity comparable across different countries and surveillance systems. The moving epidemic method (MEM) and the WHO method are widely used to this end and enable classification of season peak intensity as low, medium, high or very high. The two approaches are similar in that they base thresholds on quantiles of a normal distribution fitted to a reference set of historical observations. However, three methodological differences exist. Firstly, in the MEM the normal distribution is fitted to log-transformed incidence data, while the WHO method by default operates on the original scale. Secondly, the MEM uses more than one observation from each past season, fixing the total number to include into the reference set. The WHO method uses only the highest value from each season. Lastly, in the WHO method, but not the MEM, historical data are by default  smoothed prior to the computation of thresholds. We assess the impact of these choices on thresholds both analytically and in a simulation study. The latter is based on re-sampling of influenza-like illness (ILI) data from France and the US, thus reflecting temperate climate settings. We find that when the normal distribution is fitted to untransformed observations, a rather large proportion of new season peaks are classified as high or very high intensity. This can be mitigated by a logarithmic transformation. When fixing the total number of past observations included in the reference set as in the MEM, thresholds increase in expectation the more seasons are available. When only few are available, there is a high chance of classifying new season peaks as high or very high intensity. Smoothing incidence time series prior to computing thresholds results in somewhat less variable estimators, but also a lowering of thresholds. If these are applied to unsmoothed new season peaks, there will again be a large proportion classified as high or very high intensity. If they are applied to smoothed new data, this problem is avoided. In terms of sensitivity and positive predictive values of thresholds, we find that these cannot be expected to exceed rather modest levels if the number of available historical seasons is low; thresholds for very high intensity are particularly affected by this problem. Our practical recommendation is to include one observation per season into the reference set and employ a log transformation in the computation of thresholds. Smoothing can be applied to somewhat reduce the variability of thresholds. However, threholds then need to be applied to smoothed rather than raw new peak values, which slightly modifies their interpretation.
\end{quote}}

@@

\section{Introduction}

Influenza intensity thresholds are a common tool to make influenza activity comparable across different countries and surveillance systems. Based on data from past influenza seasons, they provide a classification into \textit{low}, \textit{medium}, \textit{high} or \textit{very high} intensity. While thresholds are designed to assess season peak intensities, they are also used to monitor spatial and temporal patterns over the course of a season. To this end, intensity classifications are commonly summarized in heat charts and maps; see Figure \ref{fig:maps} for an example from the \textit{Flu News Europe} platform run by the World Health Organization (WHO) and the European Center for Disease Prevention and Control (ECDC). Intensity thresholds thus play an important role in assessing the influenza situation at an international level.

\begin{figure}[h!]
\begin{center}
\includegraphics[width=0.6\textwidth]{figure/maps.png}

@@

\includegraphics[width=0.8\textwidth]{figure/heat_chart.png}
\end{center}
\caption{Influenza intensity map (top, week 01/2023) and heat chart (bottom, weeks 01--07/2023) as published by \textit{Flu News Europe} (\url{https://flunewseurope.org/SeasonOverview}). Copyright: World Health Organization, 2023, and European Centre for Disease Prevention and Control, 2023.
Reproduction is authorised, provided the source is acknowledged}
\label{fig:maps}
\end{figure}

Following the 2009 influenza H1N1 pandemic, the \textit{Review Committee on the Functioning of the International Health Regulations and on Pandemic Influenza (H1N1)} recommended that member states apply and evaluate their methods for severity assessment every year, thus improving their pandemic preparedness \citep[p.118]{WHO2011}. In the subsequent \textit{WHO Pandemic Infuenza Severity Assessment (PISA)} guideline \citep{WHO2017}, two statistical methods were recommended to determine influenza intensity thresholds. These are the so-called \textit{WHO method} \citep{WHO2014} and the \textit{moving epidemic method} (MEM; \citealt{Vega2015}) which has also been recommended by the European Centre for Disease Prevention and Control (e.g., \citealt{ECDC2017}). As will be detailed in Section \ref{sec:definitions}, the WHO and MEM approaches bear important similarities and can be seen as variations of the same general approach.

Especially the moving epidemic method has been adopted by many national public health agencies, and a large number of articles documenting practical applications exist (see Section \ref{sec:review}). The theoretical statistical properties of the MEM and WHO methods, however, have not yet been studied in detail. In the present work, we aim to close this gap and derive some recommendations for the computation of influena intensity thresholds. We obtain some analytical results, complemented by simulation experiments based on re-sampling of real-world data from France and the US. These indicate that using the recommended default settings, both the moving epidemic and the WHO methods tend to produce too low thresholds, leading to a higher number of season peaks classified as high or very high intensity than intended. The best behaviour of thresholds is achieved by combining characteristics of the MEM and WHO methods. Even the recommended configuration, however, is characterized by rather modest sensitivity and positive predictive values unless data on many historical seasons are available. This concerns in particular the threshold for very high intensity and should be kept in mind when using both the MEM and WHO method.

The remainder of the article is structured as follows. Section \ref{sec:desirable_properties} lists several desirable properties of a threshold setting procedure. Section \ref{sec:definitions} provides definitions of the MEM and WHO methods, highlighting three differences between these otherwise similar approaches. Section \ref{sec:review} consists of an overview of published applications of the MEM and WHO methods. This is followed by an examination of statistical properties of the two methods in Section \ref{sec:analytical_results}. In Section \ref{sec:simulation} we conduct the simulation study before concluding with a discussion in Section \ref{sec:discussion}. Here we also provide practical recommendations on the different implementation choices.

\section{Desirable properties of a thresholding method}
\label{sec:desirable_properties}

We start by stating properties of a thresholding method for influenza activity which we consider desirable. These will guide the analyses in the remainder of the paper.

\begin{description}
\item \textbf{Calibration.} As stated in the \textit{WHO Pandemic Influenza Severity Assessment} guideline, the rationale for both the MEM and the WHO method is that
\begin{quote}
\textit{``about 50--60\% of the season peaks should be above the moderate threshold, $\pm 10\%$ above the high threshold and $\pm 2.5\%$ above the extraordinary threshold''} \citep[p.10]{WHO2017}.
\end{quote}
This is a purely statistical definition, and threshold setting corresponds to estimating quantiles of an underlying distribution of season peak values. A major requirement for the thresholds is thus \textit{calibration}. We call a method calibrated if in the long run, the right fractions of seasons are classified as low, medium, high and very high. If, for instance, a method flags new season peaks as very high intensity in considerably more than 2.5\% of the cases, this will hamper the usefulness of thresholds.
\item \textbf{Sensitivity and specificity.} In addition to classifying the right number of seasons into the various categories, the individual classifications should be correct. A season peak exceeding the very high/high/medium threshold should indeed be among the 2.5\%/10\%/60\% most extreme season peaks according to the underlying true distribution. Conversely, a season peak which is not actually among the 2.5\%/10\%/60\% most extreme ones should not be flagged as such. As a complementary perspective to sensitivity and specificity, confusion matrices can be considered to assess how peaks which are truly very high, high, medium or low intensity are classified. % The true distribution of season peaks is of course unknown in practice, but the sensitivity and specificity of a given method can be studied theoretically and in simulation studies.
\item \textbf{Stability.} Estimated thresholds should not be overly variable. Ideally they would not fluctuate strongly around the corresponding quantiles of the true distribution of season peaks.
\item \textbf{Simplicity.} To ensure broad practical applicability, understanding the thresholding method should not require advanced statistical training. The method should be simple enough that a quantitatively literate user can develop a good intuition of its functioning and parameters.
\item \textbf{Ease of practical application.} Besides conceptual simplicity, ease of application is central. Methods should be straightforward to apply using well-documented and ideally open-source software packages.
\end{description}

\noindent We note that while the overarching goal of the \textit{WHO PISA} guidelines is to enhance preparedness for \textit{pandemic influenza}, the intended statistical properties of the MEM and WHO methods refer to \textit{seasonal influenza waves}. The rationale is that a good understanding of the range of peak intensities of seasonal influenza is the basis for a meaningful assessment of pandemic influenza outbreaks. Our reasoning in the following therefore focuses on the setting of subsequent waves of seasonal influenza. Note that we only address the question of peak intensity thresholds, but not season onset / baseline thresholds \citep{Vega2013}. Also, we do not address the question of determining peak intensity in real time, i.e., before the season peak has been reached or before it is clear whether a peak has been reached. Rather than simple thresholds, this would require control charts (see e.g., \citealt{Liu2019}), which we consider outside the scope of the present article.

\section{Definition of the moving epidemic and WHO methods}
\label{sec:definitions}

While framed slightly differently in the respective documentations, the MEM and WHO methods can be seen as two special cases of the same general approach. We assume that thresholds are based on weekly data (typically incidences) from $m$ past seasons and applied to the ($m + 1$)-th season. Implicitly it is further assumed that each influenza season consists of just one wave and does not feature multiple peaks separated by longer time spans. This is typically the case in temperate, but not necessarily in tropical regions. Thresholds are are then obtained as follows.

\begin{enumerate}
\item \textbf{Smoothing of historical data} (optional): apply an $l$-week moving average to all historical seasons. If data are smoothed, we denote by $x^\text{raw}_{j, k}$ the raw observation from season $j = 1, \dots, m$, week $k = 1, \dots, 52$, and by
$$
x^\text{smo}_{j, k} = \frac{1}{l} \sum_{d = 0}^{l - 1} x^\text{raw}_{j, k - d}, \ \ k = l, \dots, 52
$$ the smoothed version. In the remainder of this description we denote whichever of $x^\text{raw}_{j, k}$ and $x^\text{smo}_{j, k}$ is used to compute thresholds by $x_{j, k}$.
\item \textbf{Sorting:} Within each historical season $j = 1, \dots, m$ order all observations $x_{j, k}$ in decreasing order, denoting the $i$-th largest observation from season $j$ by $x^{(i)}_j$. More generally we will in the following use the notation $^{(\cdot)}$ to denote an ordering of values, e.g., denoting by $x^{\text{raw}, (i)}_j$ the $i$-th largest unsmoothed incidence value from season $j$.
\item \textbf{Selection of reference set:} Select the $n$ largest observations from each of the $m$ past seasons to construct a reference set $\mathcal{X} = \{x_j^{(i)}: j = 1, \dots, m; i = 1, \dots, n\}$.
\item \textbf{Data transformation:} Apply a monotonically increasing transformation $y_j^{(i)} = f(x_j^{(i)})$ to all members of the reference set $\mathcal{X}$ to obtain a reference set $\mathcal{Y}$ of transformed historical observations.
\item \textbf{Fitting a normal distribution:} Assume that the transformed values in $\mathcal{Y}$ come from a normal distribution and compute estimates of its mean and standard deviation,
\begin{align}
\begin{split}
\bar{y} & = \frac{\sum_{j = 1}^m\sum_{i = 1}^n y_j^{(i)}}{nm}\\
\sd & = \sqrt{ \frac{\sum_{j = 1}^m\sum_{i = 1}^n (y_j^{(i)}  - \bar{y})}{nm - 1}}.
\end{split}\label{eq:moments}
\end{align}
\item \textbf{Computation of thresholds:} Define intensity thresholds on the transformed scale as quantiles of the normal distribution N$(\bar{y}, \sd^2)$, i.e.\ compute
\begin{equation}
\hat{q}_{Y, \alpha} = \bar{y} + z_\alpha \sd, \label{eq:def_q}
\end{equation}
where $z_\alpha$ is the $\alpha$ quantile of the standard normal distribution. The $\hat{q}_{Y, \alpha}$ can be interpreted as estimates of quantiles $q_{Y, \alpha}$ of an underlying theoretical distribution. A common choice for $\alpha$ is
\begin{itemize}
\item the 40th percentile $\hat{q}_{Y, 0.4} = \bar{y} - 0.25 \sd$ as the threshold for medium intensity;
\item the 90th percentile $\hat{q}_{Y, 0.9} = \bar{y} + 1.28 \sd$ as the threshold for high intensity;
\item the 97.5th percentile $\hat{q}_{Y, 0.975} = \bar{y} + 1.96\sd$ as the threshold for very high intensity.
\end{itemize}
\item \textbf{Transformation of thresholds to original scale:} Obtain corresponding thresholds on the original scale by applying the inverse transformation, i.e. setting % $\hat{q}_{X, 0.4} = f^{-1}(\hat{q}_{Y, 0.4})$, $\hat{q}_{X, 0.9} = f^{-1}(\hat{q}_{Y, 0.9})$, $\hat{q}_{X, 0.975} = f^{-1}(\hat{q}_{Y, 0.975})$.
$$
\hat{q}_{X, \alpha} = f^{-1}(\hat{q}_{Y, \alpha})
$$
for $\alpha = 0.4, 0.9, 0.975$.
\item \textbf{Application of thresholds:} The obtained thresholds are applied to classify the season peak value of the $(m + 1)-th$ season. Depending on the exact specification, thresholds can be applied either to the raw peak value $x_{m + 1}^{\text{raw}, (1)}$ or the smoothed version $x_{m + 1}^{\text{smo}, (1)}$.
\end{enumerate}

\noindent The MEM and WHO methods are special cases of this procedure, see also overview in Table \ref{tab:differences}. In the MEM method, no smoothing is applied and the default data transformation $f$ is the natural logarithm.  \cite{Vega2015} recommend to use $5 \leq m \leq 10$ seasons to ensure a recent data basis. The number of observations included per season is set to $n = 30/m$, rounded to the nearest integer (with a minimum value of $n = 1$). The total number of historical observations is thus kept approximately fixed. We note that this description refers to the specification from \cite{Vega2015}, which is also reflected in the default settings of the accompanying R package \texttt{mem} \citep{Lozano2020}. The package, however, permits the user to choose $n$, $m$, and $f$ (\texttt{i.type.intensity	= 5} for no transformation, \texttt{i.type.intensity = 6} for the log transformation), thus offering considerable flexibility. We note that while \texttt{mem} does not currently allow for data smoothing using a moving average, alternatives like LOESS are available (locally estimated scatterplot smoothing; used e.g., by \citealt{Wang2023}). % The term \textit{moving epidemic method} could thus also be used as an umbrella term for the general procedure described above. We here use it in a more narrow sense for the specification from \cite{Vega2015}, reflected in the default settings of the \texttt{mem} package.
% \begin{itemize}
% \item

The implementation of the WHO method in the publicly available \textit{WHO Average Curves} application \citep{WHO2023} likewise offers a lot of flexibility; our description is based on the default settings as of May 2023, as well as the description in \cite{WHO2014}. Smoothing of data prior to the computation of thresholds is recommended, with a default of $l = 3$ (adapted from a recommendation of $l = 4$ in \citealt[p68]{WHO2014}). Subsequently, $n = 1$ observation per season is used and by default no transformation is applied to the reference set. If peak incidences vary strongly across seasons, a log transformation is recommended. At least three historical seasons are required to compute thresholds, but it is noted that the ``accuracy of these thresholds should be expected to increase with the number of seasons of good quality data available'' \cite[p22]{WHO2023}. % \citep{WHO2017}.
Thresholds are by default applied to unsmoothed new season peaks.

>>>>

\documentclass[12pt]{article}
\usepackage[left=1.5cm,right=1.5cm, top = 2cm, bottom = 2cm]{geometry}
\usepackage{graphicx}
\usepackage{natbib}
\usepackage{hyperref}
\usepackage{todonotes}
\usepackage{booktabs}
\usepackage{amssymb}
\usepackage{amsmath}

@@

\usepackage{caption}
\captionsetup[table]{name={{\footnotesize Table}}}
\usepackage{caption}
\captionsetup{font=footnotesize}

@@

% \begin{center}
% \textbf{This manuscript is is a preprint and has not been subject to peer review.}
% \end{center}

% \begin{center}
% \textbf{Running title:} Assessment of influenza intensity thresholds
% \end{center}

\abstract{
\begin{quote}
Monitoring of influenza activity is a key task of public health agencies around the world. Intensity thresholds are a commonly used tool to retrospectively classify season peak intensity and to compare current influenza activity to the peak intensity of past seasons. The resulting classifications into low, medium, high or very high intensity help to inform national risk assessment, public health planning and ressource allocation. Moreover, they feed into global risk analyses like the weekly World Health Organization (WHO) Influenza Update. Two common approaches to compute intensity thresholds, recommended in dedicated WHO guidelines, are the  % A common approach to assess season peak activity qualitatively is via intensity thresholds, which are derived empirically from peak intensities of recent seasons. to make influenza activity comparable across countries and surveillance systems. 
moving epidemic method (MEM) and the WHO method. In both methods, thresholds correspond to quantiles of a normal distribution which has been fitted to a set of potentially transformed historical observations. While an extensive body of literature on practical applications of the MEM and WHO method exists, their statistical properties have not been assessed systematically. In this paper we study the two methods analytically and in a simulation study based on re-sampling of data from French and US routine influenza surveillance. Moreover, straightforward extensions to account for small sample sizes and secular trends are described. We find that under the default settings, both the MEM and WHO method on average classify more season peaks than intended as high or very high intensity. Combining characteristics of both and adding a standard correction for small sample sizes, better-calibrated thresholds can be achieved. Even these, however, have modest sensitivity and positive predictive values. This concerns especially thresholds for the very high intensity level, which should thus be interpreted with care.
\end{quote}
}

% \begin{center}
% \textbf{Extended summary}
% \end{center}

% {\begin{quote}
% \footnotesize Intensity thresholds are a common tool to make influenza activity comparable across different countries and surveillance systems. The moving epidemic method (MEM) and the WHO method are widely used to this end and enable classification of season peak intensity as low, medium, high or very high. The two approaches are similar in that they base thresholds on quantiles of a normal distribution fitted to a reference set of historical observations. However, three methodological differences exist. Firstly, in the MEM the normal distribution is fitted to log-transformed incidence data, while the WHO method by default operates on the original scale. Secondly, the MEM uses more than one observation from each past season, fixing the total number to include into the reference set. The WHO method uses only the highest value from each season. Lastly, in the WHO method, but not the MEM, historical data are by default  smoothed prior to the computation of thresholds. We assess the impact of these choices on thresholds both analytically and in a simulation study. The latter is based on re-sampling of influenza-like illness (ILI) data from France and the US, thus reflecting temperate climate settings. We find that when the normal distribution is fitted to untransformed observations, a rather large proportion of new season peaks are classified as high or very high intensity. This can be mitigated by a logarithmic transformation. When fixing the total number of past observations included in the reference set as in the MEM, thresholds increase in expectation the more seasons are available. When only few are available, there is a high chance of classifying new season peaks as high or very high intensity. Smoothing incidence time series prior to computing thresholds results in somewhat less variable estimators, but also a lowering of thresholds. If these are applied to unsmoothed new season peaks, there will again be a large proportion classified as high or very high intensity. If they are applied to smoothed new data, this problem is avoided. In terms of sensitivity and positive predictive values of thresholds, we find that these cannot be expected to exceed rather modest levels if the number of available historical seasons is low; thresholds for very high intensity are particularly affected by this problem. Our practical recommendation is to include one observation per season into the reference set and employ a log transformation in the computation of thresholds. Smoothing can be applied to somewhat reduce the variability of thresholds. However, threholds then should be applied to smoothed rather than raw new peak values, which slightly modifies their interpretation.
% \end{quote}}

@@

\section{Introduction}
\label{sec:introduction}

Influenza is a major public health threat and monitoring of influenza activity is a central task of health authorities around the world. In order to strengthen and standardize monitoring in the aftermath of the 2009 influenza H1N1 pandemic, the World Health Organisation (WHO) developed recommendations on severity assessment during seasonal and pandemic outbreaks. According to the PISA guidelines (\textit{Pandemic Influenza Severity Assessment}, \citealt{WHO2014}), severity is defined in terms of three dimensions. \textit{Transmissibility} refers to how many people become ill and is measured e.g., via weekly case numbers of influenza-like illness (ILI) or the percentage of such cases among all consultations by general practitioners. \textit{Severity} is commonly assessed by ratios of recorded deaths and hospitalizations or hospitalizations and cases. Lastly, \textit{impact} can be measured e.g., by weekly case numbers of severe acute respiratory disease (SARI) or admissions to intensive care units.

A key role in the WHO PISA guideline is taken by influenza intensity thresholds. For transmissibility and impact indicators, they represent a generic tool to classify the peak intensity of an influenza season in light of recent peak values. Typically, classifications into \textit{low}, \textit{medium/moderate}, \textit{high} and \textit{very high/extraordinary} intensity are provided. The rationale is that
\begin{quote}
\textit{``about 50--60\% of the season peaks should be above the moderate threshold, $\pm 10\%$ above the high threshold and $\pm 2.5\%$ above the extraordinary threshold''} \citep[p.10]{WHO2017}.
\end{quote}
Similarly to the idea of a ``100-year-flood'' used to communicate hydrological risks \citep{Holmes2010}, the definition of high and very high intensity influenza seasons thus implies that on average they occur every 10 and 40 years, respectively.

Intensity thresholds form part of monitoring procedures in numerous countries, contributing to national-level risk appraisal and planning; see e.g., \cite{CDC2024} on their use in the United States. In addition, intensity classifications feed into international situation assessments by organizations like WHO or the European Center for Disease Prevention and Control (ECDC). In this context, they facilitate comparisons and summaries across countries and surveillance systems. In practice, two uses of thresholds can be distinguished \citep{CDC2024}:
\begin{itemize}
\item[(1)] In the aftermath of an influenza season, thresholds can be used to classify its peak intensity. As an example, the top panel of Figure \ref{fig:maps} shows a retrospective assessment of the 2022/23 season by the US Centers for Disease Prevention and Control (CDC; \citealt{White2023}). In terms of influenza-associated outpatient visits, hospitalization rates, and deaths among children and adolescents, this season was judged as high intensity.
\item[(2)] While an influenza season is ongoing, values of relevant indicators can be compared to intensity thresholds on a weekly basis. We note that the resulting classification is relative to past peaks rather than typical values at a given time of the season. This use of thresholds is illustrated in the bottom panel of Figure \ref{fig:maps}. In this display, published by WHO and ECDC on the \textit{FluNewsEurope} website (\url{https://flunewseurope.org/}), influenza activity across Europe is summarized in a map and a heat chart.
\end{itemize}
%As will be detailed in Section \ref{sec:definitions}, the definition and desired statistical properties of thresholds refer exclusively to season peak values. Retrospective classification of seasons is thus well-aligned with their original purpose. The use in real time, on the other hand, can be seen as a pragmatic monitoring tool, and it needs to be kept in mind that the classification is relative to past peak activity rather than usual activity at a given time of the year.
% While thresholds are designed to assess season peak intensities, they are also used to monitor spatial and temporal patterns over the course of a season. To this end, intensity classifications are commonly summarized in heat charts and maps; see Figure \ref{fig:maps} for an example from the \textit{Flu News Europe} platform run by the World Health Organization (WHO) and the European Center for Disease Prevention and Control (ECDC). Intensity thresholds thus play an important role in assessing the influenza situation at an international level.

\begin{figure}[h!]
\begin{center}
\fbox{\includegraphics[width=0.65\textwidth]{figure/mmwr_with_heading.png}}

@@

\fbox{
\includegraphics[width=0.645\textwidth]{figure/map_plus_table.png}}
\end{center}
\caption{Applications of influenza intensity thresholds by public health agencies. Top: Retrospective intensity classifications for the 2022/23 influenza season in the US by the Centers for Disease Prevention and Control (CDC; \citealt{White2023}). Here, $\text{IT}_{50}, \text{IT}_{90}$ and $\text{IT}_{98}$ denote thresholds for medium, high and very high intensity, respectively. Bottom: Influenza intensity map (week 01/2023) and heat chart (weeks 01--07/2023) as published by \textit{FluNewsEurope} (\url{https://flunewseurope.org/}). The display is collated from multiple figures and the heat chart is cropped (the original display containing an alphabetical list of all countries in the European Region of WHO). Copyright information on both figures is provided at the end of this article. % \textbf{include ``typical'' display with colours somewhere? Redondo Bravo? gov.uk?}
}
\label{fig:maps}
\end{figure}
% Following the 2009 influenza H1N1 pandemic
% The \textit{Review Committee on the Functioning of the International Health Regulations and on Pandemic Influenza (H1N1)} recommended that member states apply and evaluate their methods for severity assessment every year, thus improving their pandemic preparedness \citep[p.118]{WHO2011}.
\noindent Setting (1) is more closely aligned with the definition of thresholds in terms of nominal exceedance probabilities for new peak values, and the remainder of the article will focus on this task. We will return to setting (2) in the discussion.

In the PISA guidelines \citep{WHO2017}, two statistical methods were recommended for influenza intensity thresholding. These are the so-called \textit{WHO method} \citep{WHO2014} and the \textit{moving epidemic method} (MEM; \citealt{Vega2015}). The latter has also been recommended by ECDC (e.g., \citealt{ECDC2017}) and is used by the US CDC \citep{Biggerstaff2017}. As will be detailed in Section \ref{sec:definitions}, the WHO and MEM approaches bear important similarities and can be seen as variations of the same general approach. The two methods have been adopted by many public health agencies (see Section \ref{sec:review}). Their statistical properties, however, have not yet been studied in detail. In the present work, we aim to close this gap and derive some recommendations for the computation of influenza intensity thresholds. We study the MEM and WHO methods in their current form and suggest some extensions based on related disease monitoring methods, notably to account for small sample sizes and secular trends. We obtain some analytical results on their properties, complemented by simulation experiments based on re-sampling of routine surveillance data from France and the US. These indicate that using the recommended default settings, both the MEM and the WHO method tend to produce too low thresholds. Consequently, number of season peaks classified as high or very high intensity is higher than intended. The best behaviour of thresholds is achieved by combining characteristics of the MEM and WHO methods, along with a correction for small sample sizes. Even the recommended configuration, however, is characterized by rather modest sensitivity and positive predictive values unless data on many historical seasons are available. This concerns in particular thresholds for very high intensity, which should thus be interpreted with care. % and should be kept in mind when using both the MEM and WHO method.

The remainder of the article is structured as follows. In Section \ref{sec:desirable_properties} we list desirable properties of a threshold setting procedure, which will guide the discussion in the following. Section \ref{sec:definitions} provides definitions of the MEM and WHO methods, highlighting three differences between these otherwise similar approaches. Section \ref{sec:review} consists of an overview of published applications of the MEM and WHO methods, as well as some background on related disease monitoring methods. In Section \ref{sec:formalization_extension} we re-state the thresholding task as a statistical prediction problem and suggest some extensions to the methodology. This is followed by an examination of statistical properties of the methods in Section \ref{sec:analytical_results}. In Section \ref{sec:simulation} we conduct the simulation study before concluding with a discussion in Section \ref{sec:discussion}. Here we also provide practical recommendations on the different implementation choices.

\section{Desirable properties of intensity thresholding methods}
\label{sec:desirable_properties}

% \textbf{Add temporal structure etc here}

We start by stating properties of a thresholding method for influenza activity which we consider desirable. These will guide the analyses in the remainder of the paper. % Our arguments are based on the previously reproduced characterization of thresholds from the PISA guideline \citep{WHO2017}, which refers to the exceedance of thresholds by season peak values. % In terms of the two associated use cases, this corresponds to setting (1), i.e., the retrospective classification of season peak intensity. Setting (2), where thresholds are applied on a weekly basis over the course of a season, will only be addressed in the discussion.

\begin{description}
\item \textbf{Calibration.} As stated in Section \ref{sec:introduction}, thresholds are defined in a statistical manner -- the respective thresholds should be exceeded by 60\%, 10\% and 2.5\% of season peaks. Threshold setting thus corresponds to determining predictive quantiles for future season peak values, and a major requirement is that these are \textit{calibrated}. We call a thresholding method calibrated if in the long run, the right fractions of seasons are classified as low, medium, high and very high. If, for instance, a method flags new season peaks as very high intensity in considerably more than 2.5\% of the cases, this will hamper the usefulness of thresholds.
\item \textbf{Sensitivity and specificity.} In addition to classifying the right fraction of seasons into the various categories, the individual classifications should be correct. Assuming there is an underlying stationary distribution of season peaks, a peak which is truly among the 2.5\%/10\%/60\% highest peaks should also be flagged as such. Conversely, a peak which is not actually among the 2.5\%/10\%/60\% highest ones should not exceed the respective threshold. This aspect can only be assessed in theory and simulation studies, though; in real-world applications, there is no ``gold standard'' to determine the true intensity level of a season peak. % As a complementary perspective to sensitivity and specificity, confusion matrices can be considered to assess how peaks which are truly very high, high, medium or low intensity are classified. % The true distribution of season peaks is of course unknown in practice, but the sensitivity and specificity of a given method can be studied theoretically and in simulation studies.
\item \textbf{Stability.} Estimated thresholds should not be overly variable. Ideally they would not fluctuate strongly around the corresponding quantiles of the true distribution of season peaks.
\item \textbf{Simplicity.} To ensure broad practical applicability, understanding the thresholding method should not require advanced statistical training. The method should be simple enough that a quantitatively literate user can develop a good intuition of its functioning and parameters.
\item \textbf{Ease of practical application.} Methods should be straightforward to apply using well-documented and, ideally, open-source software packages.
\end{description}

\noindent  % while the overarching goal of the \textit{WHO PISA} guidelines is to enhance preparedness for \textit{pandemic influenza}, the intended statistical properties of the MEM and WHO methods refer to \textit{seasonal influenza waves}. The rationale is that a good understanding of the range of peak intensities of seasonal influenza is the basis for a meaningful assessment of pandemic influenza outbreaks. Our reasoning in the following therefore focuses on the setting of subsequent waves of seasonal influenza.
Note that we only address the question of peak intensity thresholds, but not baseline thresholds to determine the season onset \citep{Vega2013}; some analyses on these can be found in \cite{Pang2023}. Also, we do not address the question of classifying peak intensity in real time, i.e., before the season peak has been reached or before it is clear whether a peak has been reached. % Rather than simple thresholds, this would require control charts (see e.g., \citealt{Liu2019}), which we consider outside the scope of the present article.

\section{Definition of the MEM and WHO methods}
\label{sec:definitions}

While framed slightly differently in the respective documentations, the MEM and WHO methods are two special cases of the same general approach. We assume that thresholds are based on weekly values of a transmissibility or severity indicator (see Section \ref{sec:introduction}) from $m$ past seasons and applied to the ($m + 1$)-th season.  We denote the value for season $j$ and week $k$ by $x_{j, k}$, with more specific notation introduced below. Implicitly it is assumed that each season consists of just one wave without multiple peaks separated by longer time spans. This is typically the case in temperate, but not necessarily in tropical regions. Thresholds are then obtained via the following steps.

\begin{enumerate}
\item[(a)] \textbf{Smoothing of historical data} (optional): apply an $l$-week moving average to all historical seasons. If data are smoothed, we denote by $x^\text{raw}_{j, k}$ the raw observation from season $j = 1, \dots, m$, week $k = 1, \dots, 52$, and by
$$
x^\text{smo}_{j, k} = \frac{1}{l} \sum_{d = 0}^{l - 1} x^\text{raw}_{j, k - d}, \ \ k = l, \dots, 52
$$ the smoothed version. In the remainder of this description we denote whichever of $x^\text{raw}_{j, k}$ and $x^\text{smo}_{j, k}$ is used to compute thresholds by $x_{j, k}$.
\item[(b)] \textbf{Sorting:} Within each historical season $j = 1, \dots, m$ order all observations $x_{j, k}$ in decreasing order, denoting the $i$-th largest observation from season $j$ by $x^{(i)}_j$. More generally we will use the notation $^{(\cdot)}$ to denote an ordering of values, e.g., denoting by $x^{\text{raw}, (i)}_j$ the $i$-th largest unsmoothed incidence value from season $j$.
\item[(c)] \textbf{Selection of reference set:} Select the $n$ largest observations from each of the $m$ past seasons to construct a reference set $\mathcal{X} = \{x_j^{(i)}: j = 1, \dots, m; i = 1, \dots, n\}$.
\item[(d)] \textbf{Data transformation:} Apply a monotonically increasing transformation $y_j^{(i)} = f(x_j^{(i)})$ to all members of the reference set $\mathcal{X}$ to obtain a reference set $\mathcal{Y}$ of transformed historical observations.
\item[(e)] \textbf{Fitting a normal distribution:} Assume that the transformed values in $\mathcal{Y}$ come from a normal distribution and compute estimates of its mean and variance,
\begin{align}
\begin{split}
\bar{y} & = \frac{\sum_{j = 1}^m\sum_{i = 1}^n y_j^{(i)}}{n\times m}\\
\sd^2 & = \frac{\sum_{j = 1}^m\sum_{i = 1}^n \left(y_j^{(i)}  - \bar{y}\right)^2}{n\times m - 1}.
\end{split}\label{eq:moments}
\end{align}
\item[(f)] \textbf{Computation of thresholds:} Define intensity thresholds on the transformed scale as quantiles of the normal distribution N$(\bar{y}, \sd^2)$, i.e.\ compute
\begin{equation}
\hat{q}_{Y, \alpha} = \bar{y} + z_\alpha \sd, \label{eq:def_q}
\end{equation}
with $z_\alpha$ the $\alpha$ quantile of the standard normal distribution. The $\hat{q}_{Y, \alpha}$ can be seen as estimates of quantiles $q_{Y, \alpha}$ of an underlying distribution of peak values. The default choices are
\begin{itemize}
\item[(i)] the 40th percentile $\hat{q}_{Y, 0.4} = \bar{y} - 0.25 \sd$ as the threshold for medium intensity;
\item[(ii)] the 90th percentile $\hat{q}_{Y, 0.9} = \bar{y} + 1.28 \sd$ as the threshold for high intensity;
\item[(iii)] the 97.5th percentile $\hat{q}_{Y, 0.975} = \bar{y} + 1.96\sd$ as the threshold for very high intensity.
\end{itemize}
As we will detail in Section \ref{subsec:reformulation}, we will moreover consider an alternative formulation based on the $t$-distribution,
\begin{equation}
\hat{q}_{Y, \alpha} = \bar{y} + t_{m\times n - 1, \alpha} \times \sqrt{1 + \frac{1}{m\times n}} \times s,
\label{eq:q_Y_t}
\end{equation}
with $t_{m\times n - 1, \alpha}$ the $\alpha$ quantile of the $t$ distribution with $m\times n - 1$ degrees of freedom.
\item[(g)] \textbf{Transformation of thresholds to the original scale:} Obtain thresholds on the original scale by applying the inverse transformation, i.e. for $\alpha = 0.4, 0.9, 0.975$ set% $\hat{q}_{X, 0.4} = f^{-1}(\hat{q}_{Y, 0.4})$, $\hat{q}_{X, 0.9} = f^{-1}(\hat{q}_{Y, 0.9})$, $\hat{q}_{X, 0.975} = f^{-1}(\hat{q}_{Y, 0.975})$.
$$
\hat{q}_{X, \alpha} = f^{-1}(\hat{q}_{Y, \alpha}).
$$
\item[(h)] \textbf{Application of thresholds:} The thresholds are applied to classify the peak value of the $(m + 1)$-th season. Depending on the exact specification, thresholds can be applied either to the raw peak value $x_{m + 1}^{\text{raw}, (1)}$ or the smoothed version $x_{m + 1}^{\text{smo}, (1)}$.
\end{enumerate}
